---
title: "Yu Tian_Mid_Pstat120C"
author: "Yu Tian"
date: "2022-08-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)
```


```{r}
# package
library(tidyverse)
library(ggplot2)

# set seed
set.seed(0623)

# Read (and import) the full data set into R using read.csv()
data <- read.csv(file = 'data.csv')

# view the data example in R
data
```

## Question 1

1. Answer the following based on a simple linear regression, predicting $mpg (y)$ with $weight (x_1)$.

(a) Fit the specified model. Write the model equation, including your estimates.

#### Answer
Fit the simple linear regression model $Y = \beta_0 + \beta_1x_1 + \epsilon$ to the data.
```{r}
fit = lm(mpg~weight, data=data)
summary(fit)
```
From the table above, we can get 
$$\hat{\beta_0} = 40.2677$$
$$\hat{\beta_1} = -0.004678$$
Thus, the model equations is $$\hat{y}=\hat{\beta_0} + \hat{\beta_1}x_1+\epsilon = 40.2677 - 0.004678x_1+\epsilon$$

Using the method of least squares,
```{r}
mean_y = mean(data$mpg)
mean_y
mean_x1 = mean(data$weight)
mean_x1
```
```{r}
# n = 15
S_xy = sum(data$weight * data$mpg) - (sum(data$weight) * sum(data$mpg)/15)
S_xy
S_xx = sum((data$weight)^2) - (sum(data$weight))^2 / 15
S_xx
S_yy = sum((data$mpg)^2) - (sum(data$mpg))^2 / 15
S_yy
B_1 = S_xy / S_xx
B_1
B_0 = mean_y - B_1 * mean_x1
B_0
```

So,
$$\bar{y} = 21.9034$$
$$\bar{x_1} = 3925.809$$
$$S_{xy} = \sum_{i=1}^{n}(x_i - \bar{x})(y_i - \bar{y}) = \sum_{i=1}^{n}x_iy_i - \frac{1}{n}\sum_{i=1}^{n}x_i \sum_{i=1}^{n}y_i  = -28575.53$$
$$S_{xx} = \sum_{i=1}^{n}(x_i - \bar{x})^{2} = \sum_{i=1}^{n}x_i^2 -\frac{1}{n}(\sum_{i=1}^n x_i)^2 = 610908$$
Then, 
$$\hat{\beta_1} = \frac{S_{xy}}{S_{xx}} = \frac{8575.53}{610908} = -0.004678$$
$$\hat{\beta_0} = \bar{y} - \hat{\beta_1}\bar{x} = 40.2677$$
Thus, the model equations is $$\hat{y}=\hat{\beta_0} + \hat{\beta_1}x_1 +\epsilon = 40.2677 - 0.004678x_1 +\epsilon$$


(b) Create a scatterplot of mpg and weight. Add a line representing the model, with 95% confidence bands. Does the model appear to fit the data?

#### Answer
```{r}
data %>%
  ggplot(aes(x=weight, y=mpg)) + 
  geom_point(alpha = 0.8) + 
  labs(title = 'scatterplot of mpg and weight',
       x='weight', y='mpg') +
  geom_smooth(method = 'lm', formula = 'y ~ x', lty = 1, level=0.95) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

```
From the graph above, we can find that the 95% confidence bands covered most data points, but NOT all points are covered.
In conclusion, the model appears to fit the data with most data in confidence bands, but it does not fit all data very well. 

(c) Test the null hypothesis that the slope of $x_1, \beta_1$, is equal to zero. State the hypotheses, test statistic, rejection region(s), and p-value. Do not interpret the conclusion of this test.

#### Answer
Test of Hypothesis for $\beta_1$:
$$H_0: \beta_1 = 0$$
$$H_a: \beta_1 \neq 0$$
From the question above, we get $\hat{\beta_1} = -0.0047$, $S_{xx} = 610908$, $S_{xy}  = -28575.53$, $S_{yy} = 261.1102$. 

Then, 
```{r}
SSE = S_yy - B_1 * S_xy
SSE
S_2 = (1/(15-2)) * SSE
S_2
S_s = sqrt(S_2)
S_s

```

$$SSE = S_{yy} - \hat{\beta_1}S_{xy} = 127.4454$$
$$S^2 = (\frac{1}{n-2})\mathit{SSE} = 9.8035$$
$$s=\sqrt{S^2}=3.1311$$
Because we interested in the parameter $\beta_1$, we need the value
$$c_{11}= \frac{1}{S_{xx}} = \frac{1}{610908}$$
Then,
```{r}
t_value = B_1/(S_s*sqrt(1/S_xx))
t_value
```

$$t=\frac{\hat{\beta_1}-\beta_1}{s\sqrt{c_{11}}}=\frac{-0.0047-0}{3.1311 \cdot \sqrt{\frac{1}{610908}}}=-3.6925$$
(Besides, we could get the value of $t=-3.692$ from the table above.)

SSE is based on $df=15-2=13$. If we take $\alpha = 0.05$, the value of $t_{\alpha/2} = t_{0.025}$ for 13 df is $t_{0.025} = 2.16$, and the rejection region is 
$$\text{reject if} \ \ |t|>=2.16.$$ 
Since 3.6925 is larger than 2.16, we reject the null hypothesis that $\beta_1 = 0$.

Then, 

```{r}
p_value = 2*pt(q=-3.6925, df=13, lower.tail=TRUE)
p_value
```

(Beside, we can get the p-value = 0.00271 from the table above.)

With $t=-3.6925$, the p-value is 0.002709, which is obviously less than $\alpha = 0.05$, so we reject the null hypothesis that $\beta_1 = 0$.



## Question 2

2. Answer the following based on a multiple linear regression, predicting mpg with weight $(x1)$ and engine displacement $(x2)$.

(a) Fit the specified model. Write the model equation, including your estimates.

#### Answer
Fit the multiple linear regression model $Y = \beta_0 + \beta_1x_1 + \beta_2x_2 +\epsilon$ to the data.

```{r}
fit2 = lm(mpg~weight+displacement, data=data)
summary(fit2)
```
From the table above, we can get 
$$\hat{\beta_0} = 36.5096$$
$$\hat{\beta_1} = -0.0003083$$
$$\hat{\beta_2} = -0.07175$$
Thus, the model equations is $$\hat{y} = \beta_0 + \beta_1x_1 + \beta_2x_2 +\epsilon = 36.5095 - 0.0003083x_1 -0.07175x_2 +\epsilon$$

(b) Test the null hypothesis that the slope of $x_1$, $\beta_1$, is equal to zero. State the hypotheses, test statistic, rejection region(s), and p-value. Interpret the conclusion of this test at $\alpha = 0.05$.

#### Answer
Test of Hypothesis for $\beta_1$:
$$H_0: \beta_1 = 0$$
$$H_a: \beta_1 \neq 0$$
From the table above, we get $\hat{\beta_1} = - 0.0003083$ and the value of $t = -0.195$.

Then, since n = 15, k (the number of independent variables in the complete model) = 2, so $df$ (the degrees of freedom) $= n-k-1 = 15-2-1=12$. 

Thus, based on $df=12$. If we take $\alpha = 0.05$, the value of $t_{\alpha/2} = t_{0.025}$ for 12 df is $t_{0.025} = 2.179$. 

The rejection region is $$\text{reject if} \ \ |t|>=2.179$$ 
Since 0.195 is smaller than 2.179, we fail to reject the null hypothesis that $\beta_1 = 0$.

Then, 
```{r}
p_value = 2*pt(q=-0.195, df=12, lower.tail=TRUE)
p_value
```
(Beside, we can get the p-value = 0.849 from the table above.)

With $t=-0.195$, the p-value is 0.8487, which is obviously larger than $\alpha = 0.05$, so we fail to reject the null hypothesis and conclude that the slope is of $x_1$, $\beta_1$ is equal to zero. This means there is not a statistically significant relationship between weight and mpg.

(c) Consider $x_1^* = 3000$ and $x_2^* = 150$. Calculate a 95% confidence interval for $\mathit{E}[Y|x_1 = x_1^*, x_2 = x_2^*]$. Calculate a 95% prediction interval for $y_i$, given $x_1=x_1^*$ and $x_2=x_2^*$. Interpret both of these intervals in context.

#### Answer

* Confidence interval:
```{r}
new_data = data.frame(weight=3000,displacement=150)
confidence_interval = predict(fit2,newdata = new_data, interval='confidence', level=0.95)
confidence_interval
```

By calculation,

```{r}
# Y
Y = data$mpg
# Y'
Y_transpose = t(Y)
# X
x0 = c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1)
X = cbind(x0,data$weight,data$displacement)
# X'
X_transpose = t(X)
# the inverse of X'X
XX_inverse = solve(X_transpose%*%X)
# beta_hat = (X'X)^-(1)X'Y
beta_hat = XX_inverse %*% X_transpose %*% Y
# SSE = Y'Y-(beta_hat)'X'Y
SSE2 = Y_transpose%*%Y - t(beta_hat)%*%X_transpose%*%Y
# S^2=SSE/[n-(k+1)]
n = 15
k = 2
S2 = SSE2/(n-(k+1))
S = sqrt(S2)
# a and a'
a = c(1,3000,150)
a_transpose = t(a)
# t_(alpha/2)
t = 2.179
# confidence interval
CI1 = a_transpose%*%beta_hat + t*S*sqrt(a_transpose%*%XX_inverse%*%a)
CI2 = a_transpose%*%beta_hat - t*S*sqrt(a_transpose%*%XX_inverse%*%a)
CI1
CI2
```

Thus, a 95% confidence interval $$a'\hat{\beta}+t_{\frac{\alpha}{2}}S\sqrt{a'(X'X)^{-1}a}=27.2876$$
$$a'\hat{\beta}-t_{\frac{\alpha}{2}}S\sqrt{a'(X'X)^{-1}a}=22.3565$$


From the output (from function) above, a 95% confidence interval for $\mathit{E}[Y|x_1 = x_1^*, x_2 = x_2^*]$ the fitted mpg with (weight) $x_1^* = 3000$ and (distance)$x_2^* = 150$ is about 24.8221. The confidence interval of (22.3567, 27.2875) signifies the range in which the true mpg lies at a 95% level of confidence.

* Prediction Interval:
```{r}
prediction_interval = predict(fit2,newdata = new_data, interval='prediction', level=0.95)
prediction_interval
```

```{r}
# prediction interval
PI1 = a_transpose%*%beta_hat + t*S*sqrt(1+a_transpose%*%XX_inverse%*%a)
PI2 = a_transpose%*%beta_hat - t*S*sqrt(1+a_transpose%*%XX_inverse%*%a)
PI1
PI2
```

Thus, a 95% prediction interval $$a'\hat{\beta}+t_{\frac{\alpha}{2}}S\sqrt{1+a'(X'X)^{-1}a}=30.4394$$
$$a'\hat{\beta}-t_{\frac{\alpha}{2}}S\sqrt{1+a'(X'X)^{-1}a}=19.2048$$

From the output (of function) above, a 95% prediction interval for $y_i$, given $x_1=x_1^*$ and $x_2=x_2^*$, the fitted mpg with (weight) $x_1^* = 3000$ and (distance)$x_2^* = 150$ is about 24.8221. The prediction interval of (19.2052, 30.4389) signifies the range in which the next mpg lies at a 95% level of prediction. Notice the fitted value is the same as before, but the interval is wider. Since the prediction interval must take into account the variability of the estimators for $\mu$ and $\sigma$, the interval will be wider.


(d) Which model constitutes the “complete” model and which the “reduced” model? Can $x_2$ be dropped from the model without losing predictive information? Test at the $\alpha = 0.05$ significance level.

#### Answer


The “complete” model is (question 2(a))
$$Y = \beta_0 + \beta_1x_1 + \beta_2x_2 +\epsilon = 36.5095 - 0.0003083x_1 -0.07175x_2 +\epsilon$$
and the “reduced” model is (question 1(a))
$$Y=\hat{\beta_0} + \hat{\beta_1}x_1 +\epsilon = 40.2677 - 0.004678x_1 +\epsilon$$.

Test of Hypothesis for $\beta_2$:
$$H_0: \beta_2 = 0$$
$$H_a: \beta_2 \neq 0$$

```{r}
F_test = anova(fit, fit2)
F_test
```
From the table above, we get $SSE_R = 127.445$, $SSE_C = 64.386$, and $F=11.753$.

Thus, the F-statistic is 
$$F=\frac{(SSE_R-SSE_C)/(k-g)}{(SSE_C)/(n-(k+1))} = \frac{(127.445-64.386)/(2-1)}{(64.386)/(15-(2+1))} = 11.753$$
The tabulated F-value for $\alpha = 0.05$ with $v=2-1=1$ numerator $df$ and $v2=15-(2+1)=12$ denominator $df$ is 4.7472.

Since $11.753 > 4.7472$, so $F > F_\alpha$, which is the appropriate rejection region, so we reject the null hypothesis $H_0: \beta_2 = 0$.

From the table above, we can get the p-value is 0.005002, which is obviously smaller than $\alpha = 0.05$, so we reject the null hypothesis and conclude that the slope is of $x_2$, $\beta_2$ is not equal to zero. Thus, $x_2$ can NOT be dropped from the model without losing predictive information. This means there is a statistically significant relationship between displacement and mpg.


## Question 3

3. Consider your answers to the previous questions, then answer the following. 
Suppose that the true population relationship is given by: 
$$y = \beta_0 + \beta_1x_1 + \beta_2x_2 +\epsilon$$
Further suppose that there is a relationship between $x_1$ and $x_2$, given by:
$$x_2 = \gamma_0 + \gamma_1x_1 + \delta$$
where $\gamma_1$ and $\beta_2$ are non-zero.

(a) Find the expected values of $\beta_0$ and $\beta_1$ if the independent variable $x_2$ is omitted from the regression.

#### Answer
Given $$y = \beta_0 + \beta_1x_1 + \beta_2x_2 +\epsilon$$
Since $$x_2 = \gamma_0 + \gamma_1x_1 + \delta$$
so 
$$y = \beta_0 + \beta_1x_1 + \beta_2(\gamma_0 + \gamma_1x_1 + \delta) +\epsilon$$ 
$$= \beta_0 + \beta_1x_1 + \beta_2\gamma_0 + \beta_2\gamma_1x_1 + \beta_2\delta +\epsilon $$
$$= (\beta_0 + \beta_2\gamma_0)+ (\beta_1 + \beta_2\gamma_1)x_1 + \beta_2\delta +\epsilon$$
Thus, the expected values of $\beta_0$ is $$\mathit{E} [\beta_0] = \beta_0 + \beta_2\gamma_0$$
and the expected values of $\beta_1$ is $$\mathit{E} [\beta_1] = \beta_1 + \beta_2\gamma_1$$



(b) Calculate the bias (if any) of $\beta_0$ and $\beta_1$ when $x_2$ is omitted.

#### Answer

According to the definition of the bias,

bias of $\beta_0 = \mathit{E}[\beta_0]-\beta_0 = \beta_0 + \beta_2\gamma_0 - \beta_0 = \beta_2\gamma_0$ 

bias of $\beta_1 = \mathit{E}[\beta_1]-\beta_1 = \beta_1 + \beta_2\gamma_1 - \beta_1 = \beta_2\gamma_1$


(c) What values of $\gamma_1$ and $\beta_2$ would result in $\beta_0$ and $\beta_1$ remaining unbiased?

#### Answer

From 3(b), we get bias of $\beta_0 = \beta_2\gamma_0$ and bias of $\beta_1 = \beta_2\gamma_1$.

To make $\beta_0$ and $\beta_1$ remain unbiased, which means the bias of  $\beta_0$ and the bias of $\beta_1$ are all equal to zero.

Thus, $$\beta_2=0$$ and $$\gamma_1 = 0$$

(d) In light of the above:
i. What assumption of linear regression is being violated in Question 1? Is this assumption met in Question 2?
ii. In Question 1, are the estimates of $\beta_0$ and $\beta_1$ BLUE? Why or why not?

#### Answer

i. The assumption of $\mathit{E}[\epsilon] = 0$  is being violated in Question 1. This assumption is met in Question 2.

ii. No, the estimates of $\beta_0$ and $\beta_1$ are NOT BLUE, since NOT all assumptions are satisfied.


